---
title: "DATA 607 HW 2"
author: "Liam Byrne"
date: "September 11, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

if(!require("RMySQL", character.only = TRUE, quietly = TRUE)) {
    install.packages("RMySQL")
    library("RMySQL", character.only = TRUE)
}
```

#### Movie Review Database

The package `RMySQL` is used to access the `movieRatings` dabtabse, which was created within MySQL.

```{r}
mydb <- dbConnect(MySQL(), user = "root", password = "root", dbname = "movieratings", host = "localhost")
```

Three tables were created for the `movieRatings` database.

```{r}
dbListTables(mydb)
```

The `movies` table contains the names of the movies:

```{r}
movies <- dbGetQuery(mydb, "SELECT * FROM movies")
movies
```

The `reviewers` table stores information about the movie reviewers:

```{r}
reviewers <- dbGetQuery(mydb, "SELECT * FROM reviewers")
reviewers
```

The `ratings` table stores information about the movie ratings from the respective reviewers:

```{r}
ratings <- dbGetQuery(mydb, "SELECT * FROM ratings")
head(ratings)
```