---
title: "Data607 - HW 7"
author: "Liam Byrne"
date: "October 16, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#Reading HTML and XML objects
library("XML")
#Reading json objects
library("jsonlite")
#for getURL()
library("RCurl")
```

***

##Introduciton

The assignment envolved creating tables of our favorite books in HTML, XML and json format. A few books on my shelf that have some meanng to me are Bill Bryson's *A Short History of Nearly Everything*, Stephen Hawking's *A Brief History of Time: From the Big Bang to Black Holes* and one of my first programming books, *Data Structures and Other Objects Using C++* by Michael Main and Walter Savitch.

##Relevance

Data from the web can be in a wide array of formats, e.g., csv, HTML, XML and json. Being able to handle data from different sources and be able to coerce it into a table within R should be a fundamental skill.

***

###HTML

The 
```{r readHTML}
book_html_link <- getURL("https://raw.githubusercontent.com/Liam-O/DATA607/master/HW7/books.html")

books_html <- readHTMLTable(book_html_link,
                            header = TRUE,
                            which = 1)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
