---
title: "Data607 - HW 7"
author: "Liam Byrne"
date: "October 16, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#Reading HTML and XML objects
library("XML")
#Reading json objects
library("jsonlite")
#for getURL()
library("RCurl")
library("dplyr")
library("plyr")
```

***

##Introduciton

The assignment involved creating tables of our favorite books in HTML, XML and json format. A few books on my shelf that have some meaning to me are Bill Bryson's *A Short History of Nearly Everything*, Stephen Hawking's *A Brief History of Time: From the Big Bang to Black Holes* and one of my first programming books, *Data Structures and Other Objects Using C++* by Michael Main and Walter Savitch.

##Relevance

Data from the web can be in a wide array of formats, e.g., csv, HTML, XML and json. Being able to handle data from different sources and be able to coerce it into a table within R should be a fundamental skill.

***

###HTML

The chunk below will read the HTML file from GitHub and coerce it into a data frame object with help from `readHTMLTable` from the `XML` package.
```{r readHTML}

books_html <- readHTMLTable(
    getURL(
        "https://raw.githubusercontent.com/Liam-O/DATA607/master/HW7/books.html"),
    header = TRUE, which = 1)

class(books_html)
knitr::kable(books_html)
```

###XML

The XML is a little trickier to work with. We use `xmlToList` to coerce it into a list and read it into a data frame with the help of `ldply`, from the `XML` and `plyr` packages respectively.
```{r readXML}
books_xml <- ldply(
    xmlToList(
        getURL(
            "https://raw.githubusercontent.com/Liam-O/DATA607/master/HW7/books.xml")), data.frame) %>%
    select(-.id)

class(books_xml)
knitr::kable(books_xml)
```

##json

The json object ended up being the trickiest. It was the easiest to create an object with a subset of `authors`, but was an effort to extract the nested list to create a data table. No beautiful method could be found other than brute-forcing and casting to a list. A better solution would need to be established if the json data were more complex.
```{r}
books_json <- fromJSON(
        getURL(
            "https://raw.githubusercontent.com/Liam-O/DATA607/master/HW7/books.json"))

books_json$Author <- as.list(
    ldply(books_json$Author,
          function(x) ifelse(
            length(unlist(x))>1, paste(unlist(x), collapse = ", "),x)))
books_json <- as.data.frame(books_json)
colnames(books_json)[3] <- "Author"
knitr::kable(books_json)
```